MSF_Groupwork-NBA_shotlog

### 2016-10-12 MSFGroupwork-NBAHotHand.Rmd
enhance plot for defender distance vs accuracy

---
title: "MSF_Groupwork-NBA_shotlog" 
author: "Team_9"
output: html_document  ---
  title: "MSF_Groupwork-NBA_shotlog"
author: "Team_9"
output: html_document 
---

  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Report produced by: Justin Leiendecker, Alexander Romanenko, Artmeis Tomadaki-Balomenou, Zijun Wei, Reza Widodo  
Date submitted: 16/10/2016

@@

@@

#TO BE UPDATED when we know final areas
a. Hot Hand Theory (Artemis)
b. Defender Proximity Analysis (Reza)
c. Shot Clock Pressure  (Alex)
d. Close Game Effect - Small Margin affecting the scoring rate 
e. Home Advantage (Alex is working in this point)  
f. Fatigue Effect (Quarter Heat Rate)  
g. Shooting Percentage (James)
h. Fatigue Effect (Number of Rest Day)  
i. Regression Analysis (Alex)

## Installing libraries

```{r}
install.packages('reshape')

@@

acc <- data.frame()
acc[, "SECONDS_LEFT"] <- c(1:23)

@@

@@

game <- cast(data = df, GAME_ID1 + HOME_TEAM + AWAY_TEAM + LOCATION + player_name + W ~ FGM)

@@

```{r}

# Pick out 8 of the best players in that regular season and create a new subset. Analyse how shooting percentage of these players can influence the result of a match.

super.star.list <- c("lebron james", "james harden", "russell westbrook", "stephen curry", "anthony davis", "blake griffin", "klay thompson", "lamarcus aldridge") # players picked out
super.star <- game[game$player_name %in% super.star.list, c("player_name", "W", "FGA", "FG.percentage")] # create a subset for these players

ggplot(super.star) + geom_boxplot(aes(x = player_name, y = FGA, colour = W))
ggplot(super.star) + geom_boxplot(aes(x = player_name, y = FG.percentage, colour = W))

@@

# Correlation test
t.test(super.star[, "W"], super.star[, "FGA"]) # this makes no sense...

# Linear regression
summary(lm(W ~ player_name + FGA, data = super.star))
summary(lm(W ~ player_name + FG.percentage, data = super.star))

@@

###Defender Proximity Analysis
```{r}
#Data Preparation
#Change the value of shot result column to numeric
df[, "SHOT_RESULT"] <- as.character(df[, "SHOT_RESULT"])
for (j in 1:nrow(df)){
  df[j, "SHOT_RESULT"] <- ifelse (df[j, "SHOT_RESULT"] == "made", 1, 0)
}

@@

@@

#Create a boxplot for each player's streak lengths
boxplot(player1_df$streaks) 
summary(player1_df$streaks) #Check that boxplot is valid
IQR(player1_df$streaks) #Check that boxplot is valid
#The typical length of a streak is 0 (median = 0) - The Interquartile Range is 1 - Streak lengths of 3 and 6 are unusually high compared to the rest of the distribution
boxplot(player2_df$streaks)
summary(player2_df$streaks)
IQR(player2_df$streaks)
#The typical length of a streak is 1 (median = 1) - The Interquartile Range is 2 - There are some high streak lengths (max = 4)
boxplot(player3_df$streaks)
summary(player3_df$streaks)
IQR(player3_df$streaks)
#The typical length of a streak is 0 (median = 0) - The Interquartile Range is 0 - Streak lengths of 3 and 13 (!!!) are unusually high compared to the rest of the distribution
boxplot(player4_df$streaks)
summary(player4_df$streaks)
IQR(player4_df$streaks)
#The typical length of a streak is 1 (median = 1) - The Interquartile Range is 2 - Streak length of 6 is unusually high compared to the rest of the distribution
boxplot(player5_df$streaks)
summary(player5_df$streaks)
IQR(player5_df$streaks)
#The typical length of a streak is 1 (median = 1) - The Interquartile Range is 2 - There are some high streak lengths (max = 5)

#We can see that the distribution of all players is right skewed and that all of them (especially player3) have some long shooting streaks. To prove/disprove the hot hand theory we need to examine the indipendence of the shots. If each shot that a player takes is independent of the next shot, then a missed/made shot will not affect the propability that the next shot will be missed/made. So, if consecutive shots are independent, we disprove hot hand theory, while if consecutive shots are dependent, we prove the hot hand theory.
#Simulation - create a vector of shots('made', 'missed') for a sample shooter
set.seed(23)
shot_results <- c('made', 'missed')
simul_shooter <- sample(shot_results, size = nrow(player1_df), replace = TRUE, prob = c(0.45, 0.55))
simul_shooter

@@

#Use previous function (number_of_streaks) to calculate the streaks made by the sample shooter 
streaks_of_simulshooter <- number_of_streaks(simul_shooter)
simulshooter_df <- data.frame('streaks' = streaks_of_simulshooter)

#Comparisons of data between sample shooter and real shooters
summary(streaks_of_simulshooter)
summary(player1_df$streaks)

#Compare barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
ggplot(data = simulshooter_df, aes(x = streaks)) + geom_bar() 
ggplot(data = player1_df, aes(x = streaks)) + geom_bar()
#
```

@@

I   What would be the next steps?
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Justin Leiendecker, Alexander Romanenko, Artmeis Tomadaki-Balomenou, Zijun Wei, Reza Widodo

>>>>

---
title: "MSF_Groupwork-NBA_shotlog" 
author: "Team_9"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Report produced by: Justin Leiendecker, Alexander Romanenko, Artmeis Tomadaki-Balomenou, Zijun Wei, Reza Brianca Widodo  
Date submitted: 16/10/2016

@@

##Introduction
###What is your theme or research question?
    a. Hot Hand Theory (Artemis)
    b. Defender Proximity Analysis (Reza)
    c. Shot Clock Pressure  (Alex)
    d. Close Game Effect - Small Margin affecting the scoring rate 
    e. Home Advantage (Alex is working in this point)  
    f. Fatigue Effect (Quarter Heat Rate)  
    g. Shooting Percentage (James)
    h. Fatigue Effect (Number of Rest Day)  
    i. Regression Analysis (Alex)

@@

##Introduction
I   What is your theme or research question?
a. Hot Hand Theory (Artemis)
b. Defender Proximity Analysis (Reza)
c. Shot Clock Pressure  (Alex)
d. Close Game Effect - Small Margin affecting the scoring rate 
e. Home Advantage (Alex is working in this point)  
f. Fatigue Effect (Quarter Heat Rate)  
g. Shooting Percentage (James)
h. Fatigue Effect (Number of Rest Day)  
i. Regression Analysis (Alex)

##Theory
I   What do you expect to find?
a.

##Methods
###Dataset Description:

    Data on shots taken during the NBA 2014-2015 season for regular matches throughout the year. The data comprise of 128,609 rows and 22 columns with 16 MB data size with several value in some column is missing. The column titles are generally self-explanatory. It has the following data:   
    1.  Match ID  
    2.  Match date  
    3.  Which team plays as home and away  
    4.  Match result  
    5.  Final score difference  
    6.  Every shots attempt per game for each player  
    7.  Shot result whether it is success or not  
    8.  Type of shoot attempt (2 points or 3 points)  
    9.  Quarter number data (1 to 4)  
    10. Game clock for each quarter  
    11. Time left for a shot (shot clock)  
    12. Who took the shot  
    13. Shot distance from basket  
    14. Who was the nearest defender  
    15. Distance from the nearest defender

  The dataset is useful for evaluating who the best shooter is, who the best defender is, the hot-hand hypothesis, affect of home playing team, close game effect, shooting accuracy from best players, and defender distance effect on shooting accuracy. Data was taken from NBA's REST API provided by [Kaggle](https://www.kaggle.com/dansbecker/nba-shot-logs) and acquired at 30 September 2016.  
>>>>>>> ffa732d5e8e2f51ae1f136d3572f36becfcf2f39

###Dataset Strengths

    The dataset comprise of comprehensive observational data to start to work with. In general, the dataset only have small percentage of missing value (0.19%) so we do not need to remove or assign many data points to default values. And since the column name is not ambiguous it really helps us to understand dataset from the beginning.

###Dataset Limitations  
  There are several constraints we need to mentioned in order to provide same understandings for this group work analysis as follows:  
    1. The dataset consists only for 1 year NBA season (2014-2015)  
    2. The dataset comprise only with regular match (82 matches per team for both home and away matches)  
    3. The playoff and other matches during the year is not included
    4. The dataset does not include free throw shot and results

##Analysis

### Installing libraries

```{r}
#install.packages('reshape')

@@

acc <- data.frame()

@@

acc[, "SECONDS_LEFT"] <- c(1:23)

@@

game <- cast(data = df, GAME_ID1 + HOME_TEAM + AWAY_TEAM + LOCATION + player_name + W + FINAL_MARGIN~ FGM)

@@

```{r}
# Pick out 8 of the best players in that regular season and create a new subset. Analyse how shooting percentage of these players can influence the result of a match.

super.star.list <- c("lebron james", "james harden", "russell westbrook", "stephen curry", "anthony davis", "blake griffin", "klay thompson", "lamarcus aldridge") # players picked out
super.star <- game[game$player_name %in% super.star.list, c("player_name", "W", "LOCATION", "FINAL_MARGIN", "FGA", "FG.percentage")] # create a subset for these players

@@

ggplot(super.star) + geom_boxplot(aes(x = player_name, y = FGA, colour = W))
ggplot(super.star) + geom_boxplot(aes(x = player_name, y = FG.percentage, colour = W))
ggplot(super.star) + geom_point(aes(x = FINAL_MARGIN, y = FGA, colour = player_name))
ggplot(super.star) + geom_boxplot(aes(x = LOCATION, y = FINAL_MARGIN))

# t.test(super.star$FINAL_MARGIN[super.star$LOCATION == "H"], super.star$FINAL_MARGIN[super.star$LOCATION == "A"])

# Correlation test
cor.test(super.star$FINAL_MARGIN, super.star$FGA)
cor.test(super.star$FINAL_MARGIN, super.star$FG.percentage) # cor = 0.272

# Linear regression
summary(lm(FINAL_MARGIN ~ FGA + LOCATION, data = super.star)) # Final margin ~ FGA and location
summary(lm(FINAL_MARGIN ~ FG.percentage + LOCATION, data = super.star)) # Final margin ~ FG% and location
summary(lm(FINAL_MARGIN ~ FG.percentage + FGA + LOCATION, data= super.star))

summary(lm(FINAL_MARGIN ~ LOCATION, data = super.star)) # Final margin ~ FGA and location

@@

##Defender Proximity Analysis  
###Closest Defender Distance vs Shot Accuracy
####Introduction  
This analysis wants to calculate whether is there any significant effect from closest defender distance from a shooter with the shot outcome. The hypothesis is the further a defender from the shooter, then it will be more likely to increase the chance of scoring. This is because the shooter is practically in "free" condition without any physical interference.

###Methods  
The steps to calculate this measurement are as follows:  
  1. Assign shot result to determine whether the shot is succeed (1) or missed (0)  
  2. Select distinct of **all** closest defender distance  
  3. Count the occurence fOr every distance  
  4. Calculate how many succesfull shot scored from every distance point  
  5. Calculate accuracy by dividing successfull shot scored with distance occurence for every distance point  
  6. Plot the result
```{r}
#Calculate accuracy against defender distance
#Group the defender distance level
distance_level <- as.data.frame(table(factor(ClosestDefenderDistance)))
head(distance_level)

#Calculate how many times players scored in every defender distance
for (i in 1:nrow(distance_level)){
  distance_level[i, "SCORED"] <- sum(df[df[, "CLOSE_DEF_DIST"] == distance_level[i, "Var1"], "FGM"])
}

#Calculate shooting accuracy for each distance
distance_level$Accuracy <- round(distance_level$SCORED / distance_level$Freq, 4)

#Plotting the result
library(ggplot2)
library(scales)
ggplot(data = distance_level, aes(x = distance_level$Var1, y = distance_level$Accuracy)) +
  geom_point(aes(x = distance_level$Var1, y = distance_level$Accuracy)) +
  scale_y_continuous(labels = percent, breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0), limits = c(0, 1)) +
  scale_x_discrete("Closest Defender Distance", breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24)) +
  ylab("Shooting Accuracy (%)") +
  ggtitle("Closest Defender Distance vs Shooting Accuracy")
```
###Plot Result
  The explanation from the plot above are as follows:   
  1. If the defender stands really close to the shooter (0 to 2 feet), the accuracy varied between **42%** to **50%**  
  2. If the defender distance range from 2 (two) to 4 (four) feet, the accuracy rate for this distance range from **44%** to **49%**  
  3. If the defender distance range from 4 (four) to 6 (six) feet, the accuracy rate for this distance range from **41%** to **46%**
  4. If the defender distance range from 6 (six) to 8 (eight) feet, the accuracy rate for this distance range from **37%** to **42%**
  5. If the defender distance range from 8 (eight) to 10 (ten) feet, the accuracy rate for this distance range from **37%** to **54%**
  6. For defender distance more than 10 feet up until 16 feet, the accuracy varies greater than previous distance range, starting from **33%** to **60%**.  
  7. For the distance greater than 16 feet, the data varies between **0%** (no shots scored from that distance) to **100%** (certain successfull shot from that distance)

###Findings  
  As the analysis shown, the accuracy rate of every player in the last season was somehow **unaffected** by the defender distance. For defender distance starting from 0 (zero) to For the distance greater than 16 feet, the data varies between 0% (no shots scored from that distance) to 100% (certain successfull shot from that distance). It is also shown from the dataset that only small attempt of shots were made from further defender distance affecting the accuracy calculation result. This was arguably because further defender distance did not implied shooter distance from the basket is nearer to improve the accuracy.

###Limitations  
  As the dataset did not hold many data points for more than 20 feet defender distance, the accuracy analysis for these distances are limited. The analysis could be enhanced further by including additional season data. That way, there will be more data point for further distance and possibly would produce stronger pattern as the short distance shown.
```{r}
#Calculate accuracy against defender distance
#Group the defender distance level

distance_level <- as.data.frame(table(factor(ClosestDefenderDistance)))
head(distance_level)

#Calculate how many times players scored in every defender distance
for (i in 1:nrow(distance_level)){
  distance_level[i, "SCORED"] <- sum(df[df[, "CLOSE_DEF_DIST"] == distance_level[i, "Var1"], "FGM"])
}

#Calculate shooting accuracy for each distance
distance_level$Accuracy <- round(distance_level$SCORED / distance_level$Freq, 4)

#Plotting the result
library(ggplot2)
library(scales)
ggplot(data = distance_level, aes(x = distance_level$Var1, y = distance_level$Accuracy)) +
  geom_point(aes(x = distance_level$Var1, y = distance_level$Accuracy)) +
  scale_y_continuous(labels = percent) +
  scale_x_discrete("Closest Defender Distance", breaks=c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24, Inf)) +
  ylab("Shooting Accuracy (%)") +
  ggtitle("Closest Defender Distance vs Shooting Accuracy")

#Calculate accuracy against shooting distance
#Group the shooting distance level
shooting_distance_level <- as.data.frame(table(factor(ShotDistance)))
head(shooting_distance_level)

#Calculate how many times players scored in every shooting distance
for (i in 1:nrow(shooting_distance_level)){
  shooting_distance_level[i, "SCORED"] <- sum(df[df[, "SHOT_DIST"] == shooting_distance_level[i, "Var1"], "FGM"])
}

#Calculate shooting accuracy for each distance
shooting_distance_level$Accuracy <- round(shooting_distance_level$SCORED / shooting_distance_level$Freq, 2)

#Plotting the result
library(ggplot2)
ggplot(data = shooting_distance_level, aes(x = shooting_distance_level$Var1, y = shooting_distance_level$Accuracy)) +
  geom_point(aes(x = shooting_distance_level$Var1, y = shooting_distance_level$Accuracy))

#Data Preparation
#Change the value of shot result column to numeric
df[, "SHOT_RESULT"] <- as.character(df[, "SHOT_RESULT"])
for (j in 1:nrow(df)){
  df[j, "SHOT_RESULT"] <- ifelse (df[j, "SHOT_RESULT"] == "made", 1, 0)
}

@@

#Remove 0 from streak_per_player in order to show only the streaks achieved
streak_per_player[[1]] <- streak_per_player[[1]][streak_per_player[[1]] != 0]
for(i in 1:length(streak_per_player)) {
  streak_per_player[[i]] <- streak_per_player[[i]][streak_per_player[[i]] != 0]
}
streak_per_player

@@

#Create a boxplot for each player's streak lengths
boxplot(player1_df$streaks) 
summary(player1_df$streaks) #Check that boxplot is valid
IQR(player1_df$streaks) #Check that boxplot is valid
#The typical length of a streak is 1 (median = 1) - The Interquartile Range is 1 - Streak length of 6 is unusually high compared to the rest of the distribution
boxplot(player2_df$streaks)
summary(player2_df$streaks)
IQR(player2_df$streaks)
#The typical length of a streak is 1.5 (median = 1.5) - The Interquartile Range is 1 - Streak length of 4 is unusually high compared to the rest of the distribution
boxplot(player3_df$streaks)
summary(player3_df$streaks)
IQR(player3_df$streaks)
#The typical length of a streak is 8 (median = 8) - The Interquartile Range is 5 - There is a streak lengths of 13(!!!)
boxplot(player4_df$streaks)
summary(player4_df$streaks)
IQR(player4_df$streaks)
#The typical length of a streak is 2 (median = 2) - The Interquartile Range is 1.5 - Streak length of 6 is unusually high compared to the rest of the distribution
boxplot(player5_df$streaks)
summary(player5_df$streaks)
IQR(player5_df$streaks)
#The typical length of a streak is 2 (median = 2) - The Interquartile Range is 2 - There are some high streak lengths (max = 5)

#We can see that the distribution of all players is right skewed and that all of them (especially player3) have some long shooting streaks. To prove/disprove the hot hand theory we need to examine the indipendence of the shots. If each shot that a player takes is independent of the next shot, then a missed/made shot will not affect the propability that the next shot will be missed/made. So, if consecutive shots are independent, we disprove hot hand theory, while if consecutive shots are dependent, we prove the hot hand theory.

#Find the shooting percentage for each of the 5 players
j <- 1
shooting_percentage <- list()
for(i in best_players_df$player_id) {
  shooting_percentage[j] <- sum(df[df[ ,'player_id'] == i, 'FGM'])/ nrow(df[df[ ,'player_id'] == i, ])
  j <- j+1
}
shooting_percentage

###PLAYER 1 VS SAMPLE PLAYER###
#Simulation - create a vector of shots('made', 'missed') for a sample shooter, setting as probability the shooting percentage of the real shooter1 
set.seed(21)
shot_results <- c('made', 'missed')
simul_shooter <- sample(shot_results, size = length(streak_per_player[[1]]), replace = TRUE, prob = c(shooting_percentage[[1]], 1 - shooting_percentage[[1]]))
simul_shooter

@@

#Use previous function (number_of_streaks) to calculate the streaks made by the sample shooter 
streaks_of_simulshooter <- number_of_streaks(simul_shooter)

#Remove zeros
true_streaks_of_simulshooter <- list()
j <- 1
for(i in 1:length(streaks_of_simulshooter)) {
  if(streaks_of_simulshooter[[i]] != 0){
    true_streaks_of_simulshooter[j] <- streaks_of_simulshooter[[i]]
    j <- j + 1
  }
}
true_streaks_of_simulshooter
true_streaks_of_simulshooter <- unlist(true_streaks_of_simulshooter)

#Create a data frame 
simulshooter_df <- data.frame('streaks' = true_streaks_of_simulshooter)
simulshooter_df

#Comparisons of data between sample shooter and real shooters
summary(simulshooter_df$streaks)
summary(player1_df$streaks)

#Compare barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
ggplot(data = simulshooter_df, aes(x = streaks)) + geom_bar() 
ggplot(data = player1_df, aes(x = streaks)) + geom_bar()
#We can see that streak length distributions are quite similar, so we can conclude that Kyrie Irving doesn't have a hot hand.

###PLAYER 2 VS SAMPLE PLAYER###
#Simulation - create a vector of shots('made', 'missed') for a sample shooter, setting as probability the shooting percentage of the real shooter2 
set.seed(154)
shot_results <- c('made', 'missed')
simul_shooter <- sample(shot_results, size = length(streak_per_player[[2]]), replace = TRUE, prob = c(shooting_percentage[[2]], 1 - shooting_percentage[[2]]))
simul_shooter

#Use previous function (number_of_streaks) to calculate the streaks made by the sample shooter 
streaks_of_simulshooter <- number_of_streaks(simul_shooter)

#Remove zeros
true_streaks_of_simulshooter <- list()
j <- 1
for(i in 1:length(streaks_of_simulshooter)) {
  if(streaks_of_simulshooter[[i]] != 0){
    true_streaks_of_simulshooter[j] <- streaks_of_simulshooter[[i]]
    j <- j + 1
  }
}
true_streaks_of_simulshooter
true_streaks_of_simulshooter <- unlist(true_streaks_of_simulshooter)

#Create a data frame 
simulshooter_df <- data.frame('streaks' = true_streaks_of_simulshooter)

#Comparisons of data between sample shooter and real shooters
summary(simulshooter_df$streaks)
summary(player2_df$streaks)

#Compare barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
ggplot(data = simulshooter_df, aes(x = streaks)) + geom_bar() 
ggplot(data = player2_df, aes(x = streaks)) + geom_bar()
#We can see that the sample shooter with the same shooting percentage can have extremely high streak lengths, which means that Mo Williams does not have a hot hand.

###PLAYER 3###

best_players_df
#logistic regression

@@

##Discussion  
###What results did you find?

###Why is this interesting?

###What would be the next steps?
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
```
Justin Leiendecker, Alexander Romanenko, Artmeis Tomadaki-Balomenou, Zijun Wei, Reza Brianca Widodo
